Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis

### 2018-02-06 paper-ieee-trans.tex
Add integration testing requirement to SQA criteria

\begin{itemize}
\item \textbf{Code style}.
Code style standards are enforced for every software component in the project stack. Community
most-adopted or \textit{de-facto standard} is initially recommended, however the developer teams eventually
choose the set of guidelines to comply with. The ultimate goal for code style assessment is to
improve the readability and reusability of the source code produced under the scope of the
project.

\item \textbf{Unit and Functional testing}.
Changes involving the addition of new functionalities are requested to be tested. \textit{Regression
testing} in this context is done enforcing the successful execution of the previous
functionality tests already present in the code, so to uncover any malfunction introduced by
the changes. On the other hand, \textit{unit testing} completes the source code testing evaluation by
focusing on the code's internal design. Project's SQA criteria sets a recommended threshold of
70\% coverage for unit testing.

@@

\item \textbf{Documentation and Automated Deployment}.
Last but not least, the SQA criteria sets the path for the adoption of the developed software by
defining the documentation content, according to the target audience. These requirements also 
promote the automation, both in terms of documentation creation and service deployment. The
\textit{documentation is treated as code} so that it is portable and easily reviewed, using the same workflow
as the code does, validating the changes before being updated into the production repository. On the
other hand, the software component deployment is requested to be automated using a configuration 
management tool. These tools allow to have an \textit{infrastructure as code}, managing the component's
deployment through declarative definitions. The definition files appear as self-documenting 
code, as they sequentially guide the component's deployment process in multiple platforms.
\end{itemize}

%integrated in the Jenkins CI service \cite{indigo-jenkins}, so that any
%bug or design issue is likely to be detected and corrected promptly. The source
%code is publicly available in GitHub repositories under the project organization called
%{\sl indigo-dc} \cite{indigo-github} to increase the visibility of the products
%catalogue, promoting the external contributions and software adoption. Furthermore,
%the source code is checked for compliance with community de--facto style standards,
%selected for each one of the variety of programming languages being used as seen in
%Figure \ref{fig:fig_codestyle}.

@@

Unit testing coverage, another project quality metric, has increased between the first ({\sl INDIGO-1}) and the
second ({\sl INDIGO-2}) INDIGO--DataCloud releases, reaching an average value of
66.69\% on the {\sl INDIGO-2} release, very close to the recommended 70\% threshold
defined by the project’s SQA base criteria \cite{indigo-d31}. A comparison graph
between the code coverage for the products involved both in {\sl INDIGO-1} and
{\sl INDIGO-2} releases is shown in Figure \ref{fig:fig_unittest}. As it can be seen,
a very small set of products with low coverage values highly impacted the overall
result, although a significant part of these products are upstream contributions
or products with codebases not fully under the control of the project. Nevertheless,
60\% of the products were over this threshold while 80\% of them had values above
the 50\% coverage.

The documentation of the products is treated as source code, using a markup
language, automatically rendered and uploaded to online repositories
\cite{indigo-gitbook}. Changes in both documentation and source code are
human--reviewed as the last step before merging them into the production
repository or official documentation.

Last but not least, as described in section \ref{sec:devops}, in order
to facilitate the usage of the INDIGO--DataCloud services, they can be deployed
automatically using either Ansible \cite{indigo-ansible} or Puppet
\cite{indigo-puppet} open--source configuration tools. The adoption of such Configuration Management
tools not only eases the consumption of the products developed by external users or
infrastructures, but also tackles the correct installation and configuration of the
services and applications. A representative example is the \textit{50 roles} being 
developed within the project and currently hosted in the Ansible Galaxy portal for 
the project \footnote{https://galaxy.ansible.com/indigo-dc/}.

\begin{figure*}[ht]
\centering
\includegraphics[width=\textwidth]{images/unittest.png}
\caption{Unit testing coverage for the two major INDIGO--DataCloud releases.}
\label{fig:fig_unittest}
\end{figure*}

Figure \ref{fig:fig_confman} shows the number of products that offer an automated means
for deployment. It shows an increase after the {\sl INDIGO-1} release especially
in the previous weeks of each of the two major releases.

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.5\textwidth, height=50mm]{images/jenkins_CI_builds.png}
\caption{.}
\label{fig:fig_jenkins_CI_builds}
\end{figure}

@@

The practical implementation of the CI scenario leveraged on tightly integrated open
source tools such as GitHub \cite{github} -- as the online source code repository --,
Jenkins \cite{jenkins} -- the event-response CI system -- and Docker container
provisioning to provide instant computing power needed for executing the pipeline jobs. 
The project defined a source code contribution workflow, based on GitHub Pull Requests 
(PRs), where each change automatically triggers in Jenkins the associated quality and
metric tracking checks on PR creation or update. The set of required checks contained
in the CI pipeline report their exit status back to GitHub, preventing the change to be
approved if they are not successfully executed.

Figure \ref{fig:fig_jenkins_CI_builds}

The described CI approach is guiding the project’s software development phase throughout
the first and second major releases. Throughout its lifetime, 
% FIXME complement with the paragraph above and the references to the figures for code style and unit testing!!!!
%The implementation of a CI scenario is accomplished by automating the execution of
%the test types defined in the SQA criteria described in Section \ref{subsec:sqa}. 
%Each change in the source code will trigger the associated with the software component.
%As each change is validated, the chances of early detection of defects increases. Within
%this scenario, the cost of defect solving is dramatically reduced and the reliability of the 
%software solutions improved, as any bug or design issue is likely to be detected and subsequently
%corrected at this phase.
%

>>>>

\begin{itemize}
\item \textbf{Code style}.
Code style standards are enforced for every software component in the project stack. Community
most-adopted or \textit{de-facto standard} is initially recommended, however the developer teams eventually
choose the set of guidelines to comply with. The ultimate goal for code style assessment is to
improve the readability and reusability of the source code produced under the scope of the
project. Figure \ref{fig:fig_codestyle} showcases the code style standards and their popularity
over the software components contributing to the project.

\item \textbf{Unit and Functional testing}.
Changes involving the addition of new functionalities are requested to be tested. \textit{Regression
testing} in this context is done enforcing the successful execution of the previous
functionality tests already present in the code, so to uncover any malfunction introduced by
the changes. On the other hand, \textit{unit testing} completes the source code testing evaluation by
focusing on the code's internal design. Project's SQA criteria sets a recommended threshold of
70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
for each software component over the two major INDIGO-DataCloud releases, showing an increment in the 
averaged value of {\sl INDIGO-2} release. As it can be seen, a very small set of products with low coverage
values highly impacted the overall result, although a significant part of these products are upstream
contributions or products with codebases not fully under the control of the project. Nevertheless,
60\% of the products were over this threshold while 80\% of them had values above
the 50\% coverage.

\item \textbf{Integration testing}. Software components usually interact with other services while on
operation state. Integration testing deals with the interactions among coupled software components or
parts of a system that cooperates to achieve a given functionality. This type of testing might be 
complex and, based on the project's experience, tough to be tackled automatically. The aim is to 
guarantee the overall operation of the component, focused on the services it intefaces with, whenever
new functionalities are introduced.

@@

\item \textbf{Documentation}.
Last but not least, the SQA criteria sets the path for the adoption of the developed software by
defining the documentation content, according to the target audience. These requirements also 
promote the automation, both in terms of documentation creation and service deployment. The
\textit{documentation is treated as code}, using a markup language, automatically rendered and
uploaded to online repositories \cite{indigo-gitbook}. Thus documentation is portable and 
human--reviewed, using the same workflow as the code does, validating the changes before being
updated into the production repository.

\item \textbf{Automated Deployment}.
In order to lower the barriers of software adoption, the SQA criteria requires the automated 
deployment of the products delivered as part of the catalogue. Automation in this context is 
tackled using a configuration management tool. These tools allow to have an 
\textit{infrastructure as code} (IaC), managing the component's deployment through declarative
definitions. The definition files appear as additional source of documentation, 
self-documenting code, as they sequentially guide the component's deployment process in multiple
platforms. In this regard, INDIGO--DataCloud project contributed to open--source IaC tools such 
as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of 
such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
the Ansible Galaxy portal \footnote{https://galaxy.ansible.com/indigo-dc/}. Figure 
\ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
It shows an increase after the {\sl INDIGO-1} release especially in the previous weeks of each
of the two major releases.
\end{itemize}

@@

\subsubsection{Quality metrics}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.5\textwidth, height=50mm]{images/jenkins_CI_builds.png}
\caption{Evolution of the number of builds triggered as part of the CI implementation.}
\label{fig:fig_jenkins_CI_builds}
\end{figure}

@@

The practical implementation of the CI scenario leveraged on tightly integrated open
source tools such as GitHub \cite{github} -- as the online source code repository --,
Jenkins \cite{jenkins} -- the event-response CI system -- and Docker container
provisioning to provide instant computing power needed for executing the pipeline jobs. 
The project defined a source code contribution workflow, based on GitHub Pull Requests 
(PRs), where each change automatically triggers in Jenkins the associated quality and
metric tracking checks on PR creation or update. The set of required checks contained
in the CI pipeline report their exit status back to GitHub, preventing the change to be
approved if they are not successfully executed. As each change is validated, the chances
of early detection of defects augments. Within this scenario, the cost of defect solving
is dramatically reduced and the reliability of the software solutions improved, as any
bug or design issue is likely to be detected and subsequently corrected at this phase.

Figure \ref{fig:fig_jenkins_CI_builds} shows the evolution, throughout the 
INDIGO-DataCloud's first and second releases, of the required test types builds since the
implementation of the CI infrastructure. Automated functional testing coverage were not
available for all the software stack, thus the associated number of builds are fewer. The
executions are not precisely aligned as checks can be manually triggered, through Jenkins
interface, useful for instance whenever a failed check needs to be rebuilt.
